---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
```{r}
#creating a ruin function that can be used with changing probability, hand, repeatation times even in the american roulette problem
ruin<-function(prob, iterat, sim){
  output<-matrix(NA ,nrow = sim, ncol = 2,byrow = T) #code from Ze (will save the result in this matrix)
  colnames(output)<-c("bankroll","hands") # code from ze
for (i in 1:sim){ #loop for the replication/repeatation
  bet<-100
  bankroll<-1000
  hands<-0
while(bankroll >= bet & hands < iterat) {
   bankroll<- bankroll + (2*rbinom(1,1,prob)-1)*bet #code from Cat 
   hands<-hands + 1
 }
    output[i,1]<-bankroll
    output[i,2]<-hands
    if (hands > iterat) {break} #simulate upto nth iterations
    }
    return(output)
  }
  

```
 
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}
ruin_100<-ruin(prob=0.5, iterat=5000, sim=10000)
prob_of_bust<-function(matrix, nbet){
  busted_hands<-matrix[, "hands"][matrix[,"bankroll"]==0]
  a<-busted_hands[busted_hands<nbet]
  b<-nrow(matrix)
  prob<-length(a)/b
  return(prob)
}
prob_of_bust(ruin_100, 100)
```